# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r cache=TRUE}
# read in CSV data
data.raw <- read.csv("activity.csv", sep=",", header=TRUE)
# convert raw data to a data table
library(data.table)
data.all <- data.table(data.raw)
# convert date column to POSIXct date object
data.all$date <- as.POSIXct(strptime(data.all$date, "%Y-%m-%d"))
# convert integer interval to a 4-character 0-filled string
data.all$interval <- sprintf("%04d", data.all$interval)
head(data.all)
# identify complete cases without NAs
data <- data.all[complete.cases(data.all)]
```

## What is mean total number of steps taken per day?

```{r cache=TRUE}
# sum up daily steps
daily.steps <- tapply(data$steps, data$date, sum)
daily.steps.mean <- mean(daily.steps, na.rm=TRUE)
daily.steps.median <- median(daily.steps, na.rm=TRUE)
```

The __mean__ number of daily steps is `r sprintf("%.2f", 
daily.steps.mean)`.

The __median__ number of daily steps is `r sprintf("%.2f", 
daily.steps.median)`.

```{r cache=TRUE}
# draw histogram of average daily steps
hist(daily.steps, breaks=25, col="green",
     main="Histogram of total daily steps", 
     xlab="Total Number of Daily Steps", 
     ylab="Number of Days")
# add median line
abline(v=daily.steps.median, lty=2, lwd=2, col="black")
legend("topright", legend=c("median"), lty=2, lwd=2, col="black")
```

## What is the average daily activity pattern?

```{r cache=TRUE}
# generate daily means for each 5-minute time interval
data.interval.step.means <- tapply(data$steps, data$interval, mean)
# get the maximum number of steps for any interval during the day
max.steps <- data.interval.step.means[
    data.interval.step.means==max(data.interval.step.means)]
max.step.period <- strftime(strptime(names(max.steps), "%H%M"), format="%H:%M")
```

The __maximum__ mean number of steps during a 24-hour period was
`r sprintf("%.2f", max.steps)` at `r sprintf("%s", max.step.period)`.

```{r cache=TRUE}
# generate list of POSIX time objects of unique 5-minute intervals
# - this is used for labaling the x-axis in the line plot
x.intervals <- strptime(names(data.interval.step.means), "%H%M")
# plot the average number of steps for 5-minute intervals in a 24-hour period
plot(data.interval.step.means, type = "l", xaxt='n',
     main="Average Daily Activity Pattern",
     xlab="24 hour time of day",
     ylab="Average steps in 5-minute interval")
# create individual labels, one every hour
xlabels <- seq(x.intervals[1], x.intervals[length(x.intervals)]+300, by=3600)
# generate label locations
# - based on every 12th interval item to correspond to a 1 hour period,
#   i.e., 12 5-minute intervals = 1 hour
xat <- seq(0, length(data.interval.step.means), by=12)
# add the axis to the plot
axis(side=1, at=xat, labels=strftime(xlabels, format="%H:%M"))
# get list index (x-axis location) of the maximum value
max.steps.idx <- which.max(data.interval.step.means)
# draw line to show interval with the maximum mean number of steps
abline(v=max.steps.idx, lty=2, lwd=1, col="red")
legend.text <- paste("Max # steps at", max.step.period)
legend("topright", legend=c(legend.text), lty=2, lwd=1, col="red", cex=0.75)
```

## Imputing missing values

<!--
1. Calculate and report the total number of missing values in the dataset 
(i.e. the total number of rows with NAs)
-->

```{r}
n.rows.all <- dim(data.all)[1]
n.rows.good <- dim(data)[1]
n.nas <- n.rows.all - n.rows.good
```
The number of rows with NAs is `r n.nas`.

<!--
2. Devise a strategy for filling in all of the missing values in the 
dataset. The strategy does not need to be sophisticated. For example, you could 
use the mean/median for that day, or the mean for that 5-minute interval, etc. 
-->

```{r cache=TRUE}
# copy data with NAs to another table before modifying NAs
data.test <- data.all
# replace each instance of NA steps with the mean value for that 5-minute
# interval from all non-NA rows.
for (i in 1:nrow(data.all)) {
    if (is.na(data.all[i,]$steps)) {
        d.mean <- data.interval.step.means[data.all[i,]$interval]
        data.test[i,]$steps <- round(d.mean)
    }
}
# generate daily means for each 5-minute time interval
data.test.interval.step.means <- tapply(data.test$steps, data.test$interval, 
                                        mean)
# get the maximum number of steps for any interval during the day
max.test.steps <- data.test.interval.step.means[
    data.test.interval.step.means== max(data.test.interval.step.means)]
max.test.step.period <- strftime(strptime(names(max.test.steps), "%H%M"),
                                 format="%H-%M")
```

The __maximum__ mean number of steps during a 24-hour period was
`r sprintf("%.2f", max.test.steps)` at `r sprintf("%s", max.test.step.period)`.

```{r}
# sum up daily steps
daily.test.steps <- tapply(data.test$steps, data.test$date, sum)
daily.test.steps.mean <- mean(daily.test.steps, na.rm=TRUE)
daily.test.steps.median <- median(daily.test.steps, na.rm=TRUE)

n.test.steps.total <- sum(daily.test.steps)
n.steps.total <- sum(daily.steps)
```

The __mean__ number of daily steps is `r sprintf("%.2f", daily.test.steps.mean)`.

The __median__ number of daily steps is `r sprintf("%.2f", daily.test.steps.median)`.


```{r}
# draw histogram of average daily steps
hist(daily.test.steps, breaks=25, col="green",
     main="Histogram of total daily steps with imputed missing data", 
     xlab="Total Number of Daily Steps", 
     ylab="Number of Days")

hist(daily.steps, add=T, breaks=25, col=rgb(1, 0, 0, 0.5))
legend("topleft", c("Imputed data effect"), col=c("green"), lwd=10)

# add median line
abline(v=daily.test.steps.median, lty=2, lwd=2, col="black")
legend("topright", legend=c("median"), lty=2, lwd=2, col="black")
```


```{r}
# generate list of POSIX time objects of unique 5-minute intervals
# - this is used for labaling the x-axis in the line plot
x.test.intervals <- strptime(names(data.test.interval.step.means), "%H%M")
# plot the average number of steps for 5-minute intervals in a 24-hour period
plot(data.test.interval.step.means, type="l", xaxt='n',
     main="Average Daily Activity Pattern with Imputed Missing Data",
     xlab="24 hour time of day",
     ylab="Average steps in 5-minute interval")
# create individual labels, one every hour
xlabels <- seq(x.test.intervals[1], 
               x.test.intervals[length(x.test.intervals)]+300, by=3600)
# generate label locations
# - based on every 12th interval item to correspond to a 1 hour period,
#   i.e., 12 5-minute intervals = 1 hour
xat <- seq(0, length(data.test.interval.step.means), by=12)
# add the axis to the plot
axis(side=1, at=xat, labels=strftime(xlabels, format="%H:%M"))
# get list index (x-axis location) of the maximum value
max.test.steps.idx <- which.max(data.test.interval.step.means)
# draw line to show interval with the maximum mean number of steps
abline(v=max.test.steps.idx, lty=2, lwd=1, col="red")
legend.text <- paste("Max # steps at", max.test.step.period)
legend("topright", legend=c(legend.text), lty=2, lwd=1, col="red", cex=0.75)
```

The __maximum__ mean number of steps during a 24-hour period was
`r sprintf("%.2f", max.steps)` at `r sprintf("%s", max.step.period)`.

```{r}
delta.means <- data.interval.step.means - data.test.interval.step.means
s <- summary(delta.means)
```

Missing data imputation based on the mean value of each 5-minute period
alters the average number of steps by `r s["Mean"]` during each period.

```{r fig.width=8, fig.height=3}
par(mar=c(4,0,2,0), oma=c(1,1,0,1))
boxplot(delta.means, horizontal=TRUE,
        main="Effect of imputing missing data",
        xlab=expression(paste(Delta, " mean")))
```

## Are there differences in activity patterns between weekdays and weekends?

<!--
Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
-->

```{r}
#data.test.weekdays <- weekdays(data.test$date)
data.test$wkd <- as.factor(ifelse(weekdays(data.test$date) %in% c("Saturday",
                "Sunday"), "weekend", "weekday"))

#wkd.factor <- factor(ifelse(weekdays(data.test$date) %in% c("Saturday",
#                    "Sunday"), "weekend", "weekday"))

data.bywkd <- aggregate(data.test$steps, by=list(interval=data.test$interval, 
                                                 wkd=data.test$wkd), FUN=mean)
f <- factor(data.bywkd$wkd)

library(lattice)
xyplot(x ~ x | wkd, 
       data=data.bywkd, layout = c(1, 2))
#xyplot(data.bywkd$x ~ data.bywkd$interval, 
#       scales = list(x = list(at = 1:4, labels = levels(data.test$wkd))))

```

