# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r cache=TRUE}
# read in CSV data
data.raw <- read.csv("activity.csv", sep=",", header=TRUE)
# convert raw data to a data table
library(data.table)
data.all <- data.table(data.raw)
# convert date column to POSIXct date object
data.all$date <- as.POSIXct(strptime(data.all$date, "%Y-%m-%d"))
# convert integer interval to a 4-character 0-filled string
data.all$interval <- sprintf("%04d", data.all$interval)
head(data.all)
# identify complete cases without NAs
data <- data.all[complete.cases(data.all)]
```

## What is mean total number of steps taken per day?

```{r cache=TRUE}
# sum up daily steps
daily.steps <- tapply(data$steps, data$date, sum)
daily.steps.mean <- mean(daily.steps, na.rm=TRUE)
daily.steps.median <- median(daily.steps, na.rm=TRUE)
```

The __mean__ number of daily steps is `r sprintf("%.2f", 
daily.steps.mean)`.

The __median__ number of daily steps is `r sprintf("%.2f", 
daily.steps.median)`.

```{r cache=TRUE}
# draw histogram of average daily steps
hist(daily.steps, breaks=25, col="green",
     main="Histogram of total daily steps", 
     xlab="Total Number of Daily Steps", 
     ylab="Number of Days")
# add median line
abline(v=daily.steps.median, lty=2, lwd=2, col="black")
legend("topright", legend=c("median"), lty=2, lwd=2, col="black")
```

## What is the average daily activity pattern?

```{r cache=TRUE}
# generate daily means for each 5-minute time interval
data.interval.step.means <- tapply(data$steps, data$interval, mean)
# get the maximum number of steps for any interval during the day
max.steps <- data.interval.step.means[
    data.interval.step.means==max(data.interval.step.means)]
max.step.period <- strftime(strptime(names(max.steps), "%H%M"), format="%H:%M")
```

The __maximum__ mean number of steps during a 24-hour period was
`r sprintf("%.2f", max.steps)` at `r sprintf("%s", max.step.period)`.

```{r cache=TRUE}
# generate list of POSIX time objects of unique 5-minute intervals
# - this is used for labaling the x-axis in the line plot
x.intervals <- strptime(names(data.interval.step.means), "%H%M")
# plot the average number of steps for 5-minute intervals in a 24-hour period
plot(data.interval.step.means, type = "l", xaxt='n',
     main="Average Daily Activity Pattern",
     xlab="24 hour time of day",
     ylab="Average steps in 5-minute interval")
# create individual labels, one every hour
xlabels <- seq(x.intervals[1], x.intervals[length(x.intervals)]+300, by=3600)
# generate label locations
# - based on every 12th interval item to correspond to a 1 hour period,
#   i.e., 12 5-minute intervals = 1 hour
xat <- seq(0, length(data.interval.step.means), by=12)
# add the axis to the plot
axis(side=1, at=xat, labels=strftime(xlabels, format="%H:%M"))
# get list index (x-axis location) of the maximum value
max.steps.idx <- which.max(data.interval.step.means)
# draw line to show interval with the maximum mean number of steps
abline(v=max.steps.idx, lty=2, lwd=1, col="red")
legend.text <- paste("Max # steps at", max.step.period)
legend("topright", legend=c(legend.text), lty=2, lwd=1, col="red", cex=0.75)
```

## Imputing missing values

<!--
1. Calculate and report the total number of missing values in the dataset 
(i.e. the total number of rows with NAs)
-->

```{r}
n.rows.all <- dim(data.all)[1]
n.rows.good <- dim(data)[1]
n.nas <- n.rows.all - n.rows.good
pct.nas <- round(100 * n.nas / n.rows.all, 1)
```
The number of rows with NAs is `r n.nas` which is `r pct.nas`% of all data rows.

<!--
2. Devise a strategy for filling in all of the missing values in the 
dataset. The strategy does not need to be sophisticated. For example, you could 
use the mean/median for that day, or the mean for that 5-minute interval, etc. 
-->

```{r cache=TRUE}
# copy data with NAs to another table before modifying NAs
data.test <- data.all
# replace each instance of NA steps with the mean value for that 5-minute
# interval from all non-NA rows.
for (i in 1:nrow(data.all)) {
    if (is.na(data.all[i,]$steps)) {
        d.mean <- data.interval.step.means[data.all[i,]$interval]
        data.test[i,]$steps <- round(d.mean)
    }
}
# generate daily means for each 5-minute time interval
data.test.interval.step.means <- tapply(data.test$steps, data.test$interval, 
                                        mean)
# get the maximum number of steps for any interval during the day
max.test.steps <- data.test.interval.step.means[
    data.test.interval.step.means==max(data.test.interval.step.means)]
max.test.step.period <- strftime(strptime(names(max.test.steps), "%H%M"),
                                 format="%H-%M")
```

The __maximum__ mean number of steps during a 24-hour period was
`r sprintf("%.2f", max.test.steps)` at `r sprintf("%s", max.test.step.period)`.

```{r}
# sum up daily steps
daily.test.steps <- tapply(data.test$steps, data.test$date, sum)
daily.test.steps.mean <- mean(daily.test.steps, na.rm=TRUE)
daily.test.steps.median <- median(daily.test.steps, na.rm=TRUE)

n.test.steps.total <- sum(daily.test.steps)
n.steps.total <- sum(daily.steps)
```

The __mean__ number of daily steps is
`r sprintf("%.2f", daily.test.steps.mean)`.

The __median__ number of daily steps is
`r sprintf("%.2f", daily.test.steps.median)`.


```{r}
# draw histogram of average daily steps
hist(daily.test.steps, breaks=25, col="green",
     main="Histogram of total daily steps with imputed missing data", 
     xlab="Total Number of Daily Steps", 
     ylab="Number of Days")
# overlay histogram of data with NAs removed
hist(daily.steps, add=T, breaks=25, col=rgb(1, 0, 0, 0.5))
legend("topleft", c("Imputed data effect"), col=c("green"), lwd=10)
# add median line
abline(v=daily.test.steps.median, lty=2, lwd=2, col="black")
legend("topright", legend=c("median"), lty=2, lwd=2, col="black")
```

Replacing NAs with mean values for the interval increases the total number of
days with the mean number of steps, and the overall number of steps taken during
the complete period of the data.

<!--
```{r}
# generate list of POSIX time objects of unique 5-minute intervals
# - this is used for labaling the x-axis in the line plot
x.test.intervals <- strptime(names(data.test.interval.step.means), "%H%M")
# plot the average number of steps for 5-minute intervals in a 24-hour period
plot(data.test.interval.step.means, type="l", xaxt='n',
     main="Average Daily Activity Pattern with Imputed Missing Data",
     xlab="24 hour time of day",
     ylab="Average steps in 5-minute interval")
# create individual labels, one every hour
xlabels <- seq(x.test.intervals[1], 
               x.test.intervals[length(x.test.intervals)]+300, by=3600)
# generate label locations
# - based on every 12th interval item to correspond to a 1 hour period,
#   i.e., 12 5-minute intervals = 1 hour
xat <- seq(0, length(data.test.interval.step.means), by=12)
# add the axis to the plot
axis(side=1, at=xat, labels=strftime(xlabels, format="%H:%M"))
# get list index (x-axis location) of the maximum value
max.test.steps.idx <- which.max(data.test.interval.step.means)
# draw line to show interval with the maximum mean number of steps
abline(v=max.test.steps.idx, lty=2, lwd=1, col="red")
legend.text <- paste("Max # steps at", max.test.step.period)
legend("topright", legend=c(legend.text), lty=2, lwd=1, col="red", cex=0.75)
```
-->

Replacing NAs with mean values for the interval has minimal effect on the mean
of the total number of daily steps. The __maximum__ mean number of steps during
a 24-hour period with data imputation was
`r sprintf("%.2f", max.steps)` at `r sprintf("%s", max.step.period)`.

```{r}
delta.means <- data.interval.step.means - data.test.interval.step.means
s <- summary(delta.means)
```

Missing data imputation based on the mean value of each 5-minute period
alters the average number of steps by `r s["Mean"]` during each period.

```{r fig.width=8, fig.height=3}
par(mar=c(4,0,2,0), oma=c(1,1,0,1))
boxplot(delta.means, horizontal=TRUE,
        main="Effect of imputing missing data",
        xlab=expression(paste(Delta, " mean")))
```

## Are there differences in activity patterns between weekdays and weekends?

<!--
Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
-->

```{r}
# create a weekday vs. weekend factor
data.test$wkd <- as.factor(ifelse(weekdays(data.test$date) %in% c("Saturday",
                "Sunday"), "weekend", "weekday"))
# create a data frame dith weekday factor
data.bywkd <- aggregate(data.test$steps, by=list(interval=data.test$interval, 
                                                 wkd=data.test$wkd), FUN=mean)
# convert interval values (string at this point) to integer values
# to enable plotting with lattice xyplot.
data.bywkd$interval <- strtoi(data.bywkd$interval, 10L)
# plot weekday data with lattice xyplot
library(lattice)
xyplot(x ~ interval | wkd, type="l",
       data=data.bywkd, layout = c(1, 2),
       panel=function(x, y, ...) {
           panel.xyplot(x, y, ...)
           panel.abline(h = mean(y), lty=2) # add means
           panel.loess(x, y, degree=2, span=2/3, lty=2, col="red")
       },
       xlab="Interval", ylab="Number of Steps" )
```

```{r}
wkd.means <- tapply(data.bywkd$x, data.bywkd$wkd, mean)
wkd.max <- tapply(data.bywkd$x, data.bywkd$wkd, max)
```

The activity differs between weekends and weekdays by the mean number of steps
per interval, `r wkd.means["weekend"]` on weekends and `r wkd.means["weekday"]`
on weekdays. Weekends, therefore, tend to be more active throughout the day.

On the other hand, the maximum number of steps in a single interval, is
significantly higher during weekdays, `r wkd.max["weekday"]` vs.
`r wkd.max["weekend"]`.

###### End of report.
